Identifying the Tissue of Origin of Extracellular Vesicles Using RNA Expression Signatures

### 2015-09-11 poster/poster.Rnw
add methods section about reads include references for methods

\documentclass[a0paper,landscape]{tikzposter}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{fancybox}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{acronym}
\usepackage[backend=biber,natbib=true,hyperref=true,style=numeric-comp]{biblatex}
\addbibresource{references.bib}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
\usepackage{texshade}
\usepackage{tikz}
\usepackage{nameref}
\usepackage{zref-xr,zref-user}

@@

\begin{document}

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="0.20\\columnwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
#opts_chunk$set(cache=TRUE, autodep=TRUE)
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
library("data.table")
library("ggplot2")
library("reshape2")
library("proto")
library("xtable")
@

<<load.data,echo=FALSE,results="hide",warning=FALSE,cache.extra=tools::md5sum(paste0("../tissue_specific_expression/",c("tissue_specific_markers","interesting_gene_reads_wide")))>>=
for (a in paste0("../tissue_specific_expression/",
                 c("tissue_specific_markers",
                   "interesting_gene_reads_wide"))) {

@@

    tissue.specific.gene.reads <- 
        data.table(melt(interesting.gene.reads.wide[genes,interesting.tissues,with=FALSE],
                        id="gene_short_name",
                        variable.name="tissue",
                        value.name="FPKM"
                        ))
    tissue.specific.gene.reads[,log2.FPKM:=log2(FPKM+1)]
    return(ggplot(tissue.specific.gene.reads,
                  aes(y=gene_short_name,x=tissue))
           + geom_tile(aes(fill=log2.FPKM))
           + scale_fill_gradient(low="white",high="#002058")
           + scale_x_discrete("", expand = c(0, 0))
           + scale_y_discrete("", expand = c(0, 0))
           + theme_grey(base_size=14)
           + theme(legend.position = "none",
                   axis.ticks = element_blank(), 
                   axis.text.x = element_text(angle = 30, hjust = 1, vjust=1))
           )
}
@

\acro{EV}{Extracellular Vesicles}

@@

    One requirement for remotely sensing the transcriptome is
    identifying the source tissue of a specific \ac{EV}. To identify
    the source tissue of a specific \ac{EV}, we have identified
    tissue-specific miRNA and mRNA markers for multiple organs and
    tissues from existing publicly-available RNAseq transcriptomes by
    calculating a tissue specificity index. Using machine learning
    techniques we trained a multiclass support vector machine to
    classify the tissue of origin.

@@

  \block{Methods}{ Nothing here yet. }

@@

\block{References and Acknowledgements}{ The code and datasets for
  this poster can be found at
  \url{http://github.com/uiuc-cgm/exosome_markers_per_med_sep_2015.git}
}

>>>>

\documentclass[a0paper,landscape]{tikzposter}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{fancybox}
\usepackage{booktabs}
% \usepackage{multirow}
\usepackage{acronym}
\usepackage[backend=biber,natbib=true,hyperref=true,style=nature]{biblatex}
\addbibresource{references.bib}
% \usepackage[nomargin,inline,draft]{fixme}
% \newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
\usepackage{texshade}
\usepackage{tikz}
\usepackage{nameref}
\usepackage{zref-xr,zref-user}

@@

\begin{document}
\acrodef{EV}[EV]{Extracellular Vesicle}
\acrodef{SVM}{Support Vector Machine}

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="0.20\\columnwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
#opts_chunk$set(cache=TRUE, autodep=TRUE)
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
options(digits=2)
library("data.table")
library("ggplot2")
library("reshape2")
library("proto")
library("grid")
library("xtable")
@

<<load.data,echo=FALSE,results="hide",warning=FALSE,cache.extra=tools::md5sum(paste0("../tissue_specific_expression/",c("tissue_specific_markers","interesting_gene_reads_wide","combined_star_logs")))>>=
for (a in paste0("../tissue_specific_expression/",
                 c("tissue_specific_markers",
                   "interesting_gene_reads_wide",
                   "combined_star_logs"))) {

@@

    tissue.specific.gene.reads <- 
        data.table(melt(interesting.gene.reads.wide[genes,interesting.tissues,with=FALSE],
                        id="gene_short_name",
                        variable.name="tissue",
                        value.name="FPKM"
                        ))
    tissue.specific.gene.reads[,`log_2(FPKM)`:=log2(FPKM+1)]
    return(ggplot(tissue.specific.gene.reads,
                  aes(y=gene_short_name,x=tissue))
           + geom_tile(aes(fill=`log_2(FPKM)`))
           + scale_fill_gradient(limits=c(0,17),
                                 low="white",
                                 high="#002058",
                                 guide=guide_colourbar(title=expression(log[2](FPKM))))
           + scale_x_discrete("", expand = c(0, 0))
           + scale_y_discrete("", expand = c(0, 0))
           + theme_grey(base_size=14)
           + theme(legend.position = "top",
               axis.ticks = element_blank(), 
               plot.margin=unit(c(1,1,1,1),"mm"),
               axis.text.x = element_text(angle = 30, hjust = 1, vjust=1))
           )
}
@

@@

    One requirement for remotely sensing the transcriptome is
    identifying the source tissue of a specific \ac{EV}. To identify
    the source tissue of a specific \ac{EV}, we have identified
    tissue-specific miRNA and mRNA markers for multiple organs and
    tissues from existing publicly-available RNAseq transcriptomes by
    calculating a tissue specificity index.

    The tissue specificity index for gene $g$ ($\tau_g$) is

    \begin{equation}
      \tau_g = \frac{\sum_{i=1}^{i=N} 1-\frac{X_{i,g}}{\mathrm{max}(X_g)}}{N-1}
    \end{equation}
    \label{eq:tissue_specificity_index}

    where $N$ is the number of tissues, $X_{i,g}$ is the mean
    expression of gene $g$ in tissue $i$ and $\mathrm{max}(X_g)$ is
    the maximal expression of gene $g$ across all tissues.

    Using machine learning techniques we trained a multiclass \ac{SVM} to classify the tissue of origin.

@@

  \block{Methods}{ %
    \Sexpr{star.logs[grepl("Number of input reads",field),sum(as.numeric(value))]} %
    %
    RNAseq reads from %
    % 
    \Sexpr{length(star.logs[grepl("Number of input reads",field),value])} %
    % 
    samples corresponding to %
    \Sexpr{length(unique(na.omit(tissue.specific.genes[,tissue])))} %
    different tissues obtained from the Roadmap Epigenomics
    \cite{Bernstein.ea2010:NIHRoadmapEpigenomicsMapping} and GTEx
    \cite{Gtexconsortium.2015:Thegenotypetissueexpression} projects,
    as well as uterine (GSE50599,
    \cite{Chan.ea2014:Assessmentofmyometrialtranscriptome}) and
    placenta (GSE66622,
    \cite{Hughes.ea2015:Evaluatingintra-andinter-individual}) samples
    from the SRA.

    Reads were aligned to \textit{Homo sapiens} reference genome
    GRCh38 with Ensembl annotation release 80 using STAR v2.4.2a
    \cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and called using
    Tophat 2 v2.0.10
    \cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}. The
    average percentage of uniquely mapped reads was %
    \Sexpr{mean(as.numeric(gsub("%","",star.logs[field=="Uniquely mapped reads %",value])))}
    and the average total mapping percentage was %
    \Sexpr{mean(as.numeric(gsub("%","",star.logs[field=="Uniquely mapped reads %",value]))+as.numeric(gsub("%","",star.logs[field=="% of reads mapped to multiple loci",value])))}.
    % 
    For samples where the reads were not publicly available (GTEx and
    some Roadmap Epigenomics samples), the RPKM were taken from
    publicly available result files (see code for details).

    The \ac{SVM} was generated using caret and svmRadial with 10
    cycles of 10 repeats and a tune length of 8.

@@

\block{References}{

  \begin{center}
    \mbox{}\vspace{-\baselineskip}
  \printbibliography[heading=none]
\end{center}
}

\block{Code and Datasets}{
The code and datasets for this poster can be found at
\url{http://github.com/uiuc-cgm/exosome_markers_per_med_sep_2015.git}
